`make.traits`

### 2023-01-12 inst/gitbook/02_make.traits.Rmd
edits and cleanings

@@

For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):

```{r}
## A valid (but useless) process
valid.process <- function(x0, egde.length = 1, ...) {
    return(42)
}
```

> Note that the argument `edge.length` is set to `1` by default. In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speeding up the `dads` function).

@@

```{r}
## A wrongly formated process
invalid.process <- function(x0) {
    return(rnorm(1, mean = x0))
}
```

The `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.

@@

Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using:

@@

@@

This bit is more for development.
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

@@

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `dads` processes to set all the arguments but `x0` with default values (just in case).
Also, note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but are rather generalised/simplified version that you can use as a template

@@

```{r}
## A simple Brownian motion process
my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
    ## Drawing a random number from a normal distribution
    ## with x0 as the and a given standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = x0, sd = sd * edge.length)

@@

```{r}
## A simple Ornstein-Uhlenbeck motion process
my.OU.process <- function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
    ## Calculate the mean based on alpha
    mean <- x0 * exp(-alpha)
    ## Calculate the standard deviation based on alpha and the variance
    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
    ## Draw a random number from a normal distribution
    ## using this mean and standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = mean, sd = sd * edge.length)

>>>>

### The syntax (how to code a process?)

@@

For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
In this example, the process is not dependent on neither the previous state (`x0`) and the branch length (`edge.length`).

```{r}
## A valid (but useless?) process
valid.process <- function(x0 = 0, edge.length = 1, ...) {
    return(42)
}
```

> Note that in this function definition the arguments `x0` and `edge.length` have a default value set to `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

@@

```{r}
## A wrongly formated process
invalid.process <- function(x0 = 0) {
    return(rnorm(1, mean = x0))
}
```

### Using a `"process"` in `dads`

You can design your own process as a function (as long as it has a valid syntax).
Alternatively, the `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.

@@

Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `dads` package):

@@

### What even is a trait?

The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
<!--TODO: add REF e.g.  https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.8321 -->
Because it would be impossible to accommodate all definitions in `dads` we had to go with an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc...) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
In `dads`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `dads`:

```{r, eval = FALSE}
## Three traits with one process:
make.traits(BM.process, n = 3, start = c(1,2,3))
## Six traits with two processes:
make.traits(c(BM.process, OU.process), n = 3)
## Five traits with two processes
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

@@

**This bit is more for development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

@@

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `dads` processes to set all the arguments with default values (just in case).

> Note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but an easier to edit version that you can use as a template:

@@

```{r}
## A simple Brownian motion process
my.BM.process <- function(x0 = 0, edge.length = 1, sd = 1, ...) {
    ## Drawing a random number from a normal distribution
    ## with x0 as the and a given standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = x0, sd = sqrt(sd^2 * edge.length))

@@

```{r}
## A simple Ornstein-Uhlenbeck motion process
my.OU.process <- function(x0 = 0, edge.length = 1, var = 1, alpha = 1, ...) {
    ## Calculate the mean based on alpha
    mean <- x0 * exp(-alpha)
    ## Calculate the standard deviation based on alpha and the variance
    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
    ## Draw a random number from a normal distribution
    ## using this mean and standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = mean, sd = sqrt(sd^2 * edge.length))
